PAPER DRAFT

### 2021-11-29 Papers/Polymethylation_scores/02_abstract.Rmd
comments from erin

**Background**: Prenatal maternal smoking has negative implications for child health.  DNA methylation signatures  can function as a biomarker of prenatal maternal smoking. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues. In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores. We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using Area Under the Curve (AUC) from receiver operating curve analyses.

**Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score. The direction and magnitude of the association was consistent across when stratified genetic ancestries. In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001), and more accurate than using a single *a priori* site in the AHRR gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).

>>>>

**Background**: Prenatal maternal smoking has negative implications for child health.  DNA methylation signatures can function as biomarkers of prenatal maternal smoking. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues. In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores for prenatal maternal smoke exposure. We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses.

**Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score. The direction and magnitude of the association was consistent when stratified by genetic ancestries. In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001). The polymethylation scores showed higher classification accuracy than using a single *a priori* site in the *AHRR* gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).

### 2021-11-29 Papers/Polymethylation_scores/04_background.Rmd
comments from erin

Maternal prenatal smoking is a public health concern. In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard for smoking measurements, serum cotinine levels, have a short half-life of nine hours in pregnant women [@Dempsey:2002wr]. Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical sample, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.

Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across genetic ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of the associations between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.

After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. Biomarkers can consist of single DNA methylation sites or summary meausres across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. One attractive option is a polymethylation score, analogous to a polygenic score. In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. One such DNA methylation score calculated from cord blood classified maternal smoking behavior with an area under the curve (AUC) of 0.82 in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior was classified with an AUC of 0.87 [@Ladd-Acosta:2016aa]. Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at DNA methylation sampling and portable across genetic ancestry groups. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age, and ancestry has not been evaluated in the same study.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoking. Before translation to clinical applications, we must understand how and if the signal varies across age, ancestry, and tissue sample. In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites. We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.

>>>>

Maternal prenatal smoking is a public health concern. Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard for smoking measurements, serum cotinine levels, have a half-life of nine hours in pregnant women, limiting the ability for accurate detection to a short window [@Dempsey:2002wr]. Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.

In infant cord blood and placental tissues, maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, though this has primarily been tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation outside of cord blood and placental tissue. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical sample, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.

Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across genetic ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of the associations between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom 28 reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.

After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. Biomarkers can consist of single DNA methylation sites or summary measures across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. One option is a polymethylation score, analogous to a polygenic score. In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. One such DNA methylation score calculated from cord blood classified maternal smoking behavior with an area under the curve (AUC) of 0.82 in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior was classified with an AUC of 0.87 [@Ladd-Acosta:2016aa]. Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. Ideally, a biomarker for prenatal maternal smoking would be consistent over the life course and portable across genetic ancestry groups. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age, tissue, and genetic ancestry has not been evaluated in the same study.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoking. Before translation to clinical applications, we must understand how and if the signal varies across age, tissue sample, and genetic ancestry. In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual *a priori* CpG sites, global DNA methylation, and epigenetic clocks. We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.
